An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-02-07 rule-ltl.tex
Updates from Overleaf

Linear Temporal Logic (LTL) is a common specification formalism for reactive and concurrent systems. It is often used in e.g. model checking and runtime verification. Another formalism that is used for the same purpose is finite automata, often over infinite words. This includes B\"{u}chi, Rabin, Street, Muller and Parity automata~\cite{Thomas}, all having the same expressive power. In fact, model checking of an LTL specification
is usually performed by first translating the specification into a B\"{u}chi automaton. The automata formalisms are more expressive than LTL, with a classical example By Wolper~\cite{Wolper} showing that
it is not possible to express in LTL that every even state in
the sequence satisfies $p$. This has motivated extending LTL
in various ways.
We are mainly motivated by runtime verification of executions with data, for which a first-order temporal logic is
appropriate. There, too, we show that expressiveness can be extended.

We present first an extension of the propositional LTL that is based on using additional (auxiliary) propositions, not appearing in the execution, which summarize the past 
of the current execution. These propositions obtain their value in a state as a function of the prefix of the execution up and including that state through a past temporal formula. This can be seen as adding
summary states to the temporal specification, resulting in a specification that is in between logic and finite automata representation. This extension, which we believe is conceptually simpler
and more intuitive than other suggested extensions, is shown to be
equivalent to the other common specification formalism:
B\"{u}chi automata, regular expressions and quantified LTL.

We then proceed to first-order LTL. We relativise Wolper's example to the first
order case to motivate the need for
extending first-order LTL. Another example is based on the inexpressiveness of transitive closure of a relation in first order logic.
We present an extension where auxiliary relations, not appearing in the model, are
added. While this extension
is capable of expressing the above
properties, we show that for the first-order version, this has less 
expressive power than adding quantification over relations, in distinction from the corresponding propositional case.

These extensions to the logics are in particular natural for
runtime verification of past (safety) temporal properties, since
the values of propositional and first-order LTL are based
on calculating and updating a summary of the prefix of the execution. Runtime verification often concentrates on a past  version of temporal logic~\cite{HR} mainly because for these properties we can decide when the monitored property is violated immediately when this happen, while for future properties we may never know for sure. Consistent with that, 
we provide runtime verification algorithms for the past-time versions of the extended propositional and first-order
logics. We further present an extension of the 
\dejavu{} tool (https://github.com/havelund/dejavu) that
realizes the extended first-order past temporal logic verification. The \dejavu{} tool~\cite{HPU,HP} allows runtime verification on past time first order temporal logic over infinite domains (e.g., the integers, strings). It achieves efficiency by BDD representation of enumerations of data (rather than over the data values themselves) and using
a garbage collection algorithm. 
Finally, we provide some experimental results.

The contributions of this paper are both theoretical and practical. On the theory side, we present and study extensions for propositional and first order linear-temporal logics and show the relations between them and existing versions. 
On the practical side, we present algorithms for the extended logic, and 
show how to cast them specifically in an efficient BDD-based implementation. We present some experiments performed with our implementation and report on the results.

The structure of this paper is as follows. In Section~2 we first define the syntax and semantics of propositional temporal logic. We then review some of its classical extensions and present our own extension. In Section~3 we present a runtime verification algorithm for the extended past LTL. In Section~4 we review first order linear temporal logic, present our extension and study the relationships between the presented versions of first-order LTL. IIn Section~5 we present a runtime verification algorithm for the past version of the extended first-order LTL. Section~6 describes the implementation and provides experimental results. Finally, Section~7 concludes the paper.

@@

\vspace{1ex}
\noindent {\bf Related Work.}
For propositional LTL, the logic ETL~\cite{WVS,Wolper} extends
the set of temporal operators 
by using automata (or, equivalently a right linear grammar). Each automaton defines a temporal operator,
where the input letters of the automaton correspond to future temporal subformulas. Each accepted run defines then the places
where the subformulas (corresponding to the occurrences of the input letters in the run) need to hold. Another expressive-equivalent extension, 
in~\cite{Wolper}, adds
state-dependent quantification over propositions to LTL.

@@

\begin{itemize}
\item{P} restricts to P{\em ast} temporal operators.
\item{F} allows F{\em irst-order} quantification over
data assigned to variables that is uniform for all states. This quantification is referred to as {\em static} quantification.
\item{Q} adds (second order) Q{\em uantification} that is state-dependent. This is referred to as {\em dynamic} quantification.
\item{E} allows E{\em xtending} the objects that occur in the input with
new auxiliary predicates (for the propositional versions) or relations (for the first-order cases) and past temporal logic rules that define them.
\end{itemize}

@@

\section{Propositional LTL}

The classical definition of linear temporal logic~\cite{MP} has the following syntax:
\[ \varphi ::= \true \, | \,  p \, | \, ( \varphi \wedge \varphi ) \, |  \, \neg  \varphi \, |   \, \bigcirc \varphi \, |  
\, ( \varphi \until \varphi ) \,|  \, \ominus \varphi |
   \, ( \varphi \since \psi ) \,  \]
where $p$ is a proposition from a finite set of propositions $P$, and with $\bigcirc$, $\until$, $\ominus$, $\since$ standing for {\em next-time}, 
{\em until}, {\em previous-time} and {\em since}, respectively.

The models for LTL formulas 
are infinite sequence of states, of the form
$\sigma =  s_{1}\, s_{2}\, s_3 \ldots$,
where $s_i \subseteq P$ for each $i\geq 0$. These are the propositions that {\em hold} in that state.
We denote by $\xi_{i}$ the suffix $s_{i} s_{i+1} s_{i+2}\ldots$ of $\sigma$. 
LTL's semantics is defined as follows:
\begin{itemize}
\item $( \sigma, {i}) \models \mathit{true}$.
\item $(\sigma, {i}) \models p$ iff $p \in s_{i}$.
\item $(\sigma, {i}) \models \neg\varphi$ iff  $(\sigma, {i}) \not\models \varphi$.
\item $(\sigma, {i}) \models ( \varphi \land \psi)$ iff $(\sigma, {i}) \models \varphi$ and $(\sigma, {i} ) \models \psi$.
\item $(\sigma, {i}) \models \bigcirc \varphi$ iff $(\sigma, {i+1}) \models \varphi$.
\item $(\sigma, {i}) \models ( \varphi \cal{U} \psi )$ iff for some $j$,
$j\geq i$, $(\sigma, {j}) \models \psi$,
and for each $k$, $i \leq k < j$, $(\sigma, {k}) \models \varphi$.
\item $(\sigma , i) \models \ominus \varphi$ iff $| \sigma | > 1$ and $(\sigma, i-1) \models \varphi$.
\item $(\sigma , i) \models (\varphi \since \psi)$ iff there exists $j$, $1 \leq j \leq i$, such that
$(\sigma , j)  \models \psi$ and for each
$k$, $j < k \leq i$, $(\sigma , k ) \models \varphi$.
\end{itemize}
Then $\sigma \models \varphi$ when $( \sigma ,0 ) \models \varphi$.
We can use the following abbreviations:
$\false = \neg \true$, 
$(\varphi \vee \psi) = \neg ( \neg \varphi \wedge \neg \psi )$, 
$(\varphi \rightarrow \psi) = ( \neg \varphi \vee \psi)$,
$(\varphi \leftrightarrow \psi) = ((\neg \varphi \wedge \neg \psi) \vee (\varphi \wedge \psi ))$,
$\Diamond \varphi = (\true \until \varphi)$ and $\Box \varphi = \neg \Diamond \neg \varphi$,
${\bf P}\  \varphi = (\true \; \since \; \varphi)$ ({\bf P} stands for P{\em reviously}), and 
${\bf H}\  \varphi = \neg {\bf P}\ \neg \varphi$.

%%infinite sequences (finite executions can be extended by repeating the last event indefinitely to form an infinite paths).
The expressive power of propositional LTL is often compared to regular expressions
over the alphabet $\Sigma = 2^P$ and to  {\em monadic} first and second order logics where the temporal propositions
correspond to unary predicates (over time). An overview appears in~\cite{Thomas}. Accordingly, we have the following characterizations:
LTL is equivalent to monadic first-order logic\footnote{First-order logic with unary predicates over the naturals, and the relation $<$.},  star-free regular expressions\footnote{Regular expressions but without star (or $\omega$-star for infinite words).} and counter-free B\"{u}chi automata\footnote{A counter-free language is a regular language for which there is an integer $n$ such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}. In fact, restricting the temporal operators to the {\em future}  operators $\until$ and $\bigcirc$
(and the ones derived from them $\Box$ and $\Diamond$)
leaves the same expressive power and is commonly used.

@@

An important subset of LTL, called here PLTL, allows only past temporal
operators: $\since$, $\ominus$ and the ones derived
from them, $\mathbf{H}$ and $\mathbf{P}$. The past time logic is often
(and also here) interpreted over finite sequences, 
where $\sigma \models \varphi$ when $( \sigma , | \sigma | ) \models \varphi$.
Alternatively,
it is also a common practice to use the past time restricted version, prefixed with  a single $\Box$ (always) operator; there, each of the prefixes has to satisfy $\varphi$. This later
form expresses {\em safety} LTL properties~\cite{AS} over infinite sequences.

When PLTL is interpreted over finite sequences, its
expressive power is the same as star-free regular expressions, first-order monadic logic over finite sequences and counting-free automata.

@@

\subsubsection{Extending LTL with Dynamic Quantification.}
Adding quantification over propositions allows writing a formula of the form
$\Exists q \, \varphi$, where
$\Exists q$ repesents {\em existential} quantification over a propositiona $q$ that can appear in $\varphi$.

@@

\begin{itemize}
\item $( \sigma, i ) \models \boldsymbol{\Exists} q \, \varphi$ iff there exists
$\sigma'$ such that $\sigma' |_{\{ q \} } = \sigma$ and 
$( \sigma', i) \models \varphi$.
\end{itemize}
%Since all the auxiliary variables appear in the scope of
%quantifiers, it holds that $\sigma \models \varphi$
%(in $\varphi$, auxiliary variables appear only within
%the scope of quantifiers)
%iff $\sigma |_A \models \varphi$.
%
%\klaus{Are the last two lines above needed?}
%
Universal quantification is also allowed, where $\Forall q\, \varphi =
\neg \Exists q\, \neg \varphi$.
This kind of quantification is considered to be ``dynamic'', since the quantified variables can have different truth values depending on the states It is also callsed  "second order" quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
Extending LTL with such quantification, the logic QLTL has the same expressive power as regular expressions (over infinite words), full B\"{u}chi
automata, or monadic second order logic with unary predicates over the naturals, see again~\cite{Thomas}.
In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of QLTL~\cite{Thomas}.

\noindent
Wolper's property can be expressed in QLTL as follows:
\begin{equation} \label{form0} \Exists q \,
( (  \neg q \wedge \Box (q \leftrightarrow \bigcirc \neg q )  )  \wedge \Box (q \rightarrow p ) )\end{equation}

If we allow
quantification over propositions, we obtain the logic PQLTL
(QLTL restricted to the past modalities), with the following formula:
\begin{equation} \label{form1} \Exists q \,
{\bf H} (( q \leftrightarrow \ominus \neg q )   \wedge (q \rightarrow p ) )\end{equation}
%In fact, it can be shown that a single existential quantifier is sufficient. 
PQLTL has the same expressive
power as regular expressions and finite automata.
Note that since finite automata over finite sequences can be determinized, it is always sufficient to translate a
PQLTL property into a deterministic
automaton.

@@

We present now an extension of LTL, which we call ELTL. It has the same
expressive power as  QLTL.
It has a form that puts it, conceptually, in between temporal
logic and automata: on the one hand, it is possible to simply write
a pure LTL formula, and on the other hand, one can write a specification that resembles a finite automaton. As will be showed later, this extension is
is very natural for runtime verification.

We partition the propositions $P$ into
{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
and {\em basic propositions} $B$. A model for an ELTL formula has truth assignments only over the propositions in $B$.
An ELTL property $\eta$ has the following form: 
\begin{equation} \label{ELTL}
\psi \mathit{\ where\ } a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}} \end{equation}
where each $a_i$ is a distinct auxiliary proposition from $A$,
$\psi$ is an LTL property and each $\varphi_i$ 
is a PLTL property where propositions from $B$ can
only occur within the scope of a $\ominus$ operator.
The semantics can be defined as follows. We refer to $\psi$
as the {\em statement} of $\eta$ and to 
$a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}$ as the {\em rules}. In text, rules
 will be separated from each other by a comma (",").

\begin{tabbing}
\ \ \ $\sigma \models \eta$ \=  if there exists $\sigma'$, where
$\sigma' |_A = \sigma$ such that \\
 \> $\sigma' \models ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j)) $
\end{tabbing}

\noindent
Wolper's example can be written in ELTL
as follows:
\begin{equation} 
\label{form3}
\Box ( q \rightarrow p ) \, \mathrm{where} \, 
 ( q := \ominus \neg q)
\end{equation}
where $A = \{ q\}$ and $B = \{ p \}$.
The auxiliary variable $q$ is used to augment the input sequence such that each {\em odd} state will satisfy $\neg q$
and each {\em even} state will satisfy $p$.  As $\ominus \varphi \models \false$ in the initial state, regardless of $\varphi$, then
$q$ is $\false$ in the initial state, and then it subsequently alternates.

ELTL extends the set of propositional variables with new variables whose values at a state are
based on the prefix of the model. The added propositions are dependent and effectively summarize the past execution prefix up to and
including the current state.
This differs from 
the use of auxiliary variables in QLTL, where the auxiliary variables do not have to extend the execution
states in a fixed way throughout the interpretation of 
the property over a model. The constraint that auxiliary variables appearing in the formulas $\varphi_i$ 
must occur within the scope of a $\ominus$ operator
is required to help preventing that the values
of the auxiliary variables are not conflicting, as in
$a_1 := \neg a_2$ and $a_2 := a_1$, which makes a formula
that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write conflicting rules, e.g., by simlpy 
adding the rule $a: = \false$.

@@

\noindent {\bf Sketch of proof.} By induction on the state index. Each auxiliary variable
is defined, via a PLTL formula by the current state of the model over the values of the propositions $B$, and by the values of $A \cup B$ in previous states. A PLTL
property depends in a state only on the prefix
of the model that precedes and includes that state. \qed

\begin{theorem}
\label{theo1}
The expressive power of 
ELTL is the same as QLTL.
\end{theorem}
\noindent {\bf Proof.}
Each ELTL formula $\eta$, as defined above,
is expressible in QLTL using the following formula:

@@

For the other direction, one can translate the QLTL property first into
a second order monadic logic formula, then to a deterministic Muller automata and then constructs an ELTL formula that holds for the accepting executions of this automaton. The details of the
constructions are given in the appendix. \qed

@@

We extend PLTL into EPLTL by disallowing the future temporal operators (except an implied prefixing $\Box$). This results in
a formalism that is equivalent to finite automata, where all the states except one are accepting and where the nonaccepting state is a sink. 
We can use related, but simpler construction than in Theorem~\ref{theo1} to prove the following:

@@

\begin{lemma}
\label{simplecase}
The expressive power of 
ELTL is the same as QLTL.
%Any safety QPLTL property can be expressed using the following restricted
%ELTL form:
%\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
% {j \in \{1, \ldots , n\}}  \] 
% with $p$ being one of the auxiliary variables %$a_j$.
\end{lemma}
In fact, it is easy to show that EPLTL can be restricted to the following form, while retaining its expressive power.
\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
{j \in \{1, \ldots , n\}}  \] 
with $p$ being one of the auxiliary variables $a_j$. The relative expressive power between the propositional temporal logics discussed in this section is summarized in Figure~\ref{prop}.

@@

\begin{figure}
\begin{center}
%\includegraphics[height=0.6in,width=2in]{PROP.jpg}
\includegraphics[width=3.5in]{PROP.jpg}
\caption{\label{prop} An arrow from A to B means that logic  B is more expressive than A. A double headed arrow shows same expressiveness}
\end{center}
\end{figure}

@@

Runtime verification of temporal specification often
concentrates on the past portion of the logic.  Past-time  specifications have the important property that one can distinguish when they are violated. For an extended discussion of this issue of {\em monitorability}, see e.g.,~\cite{Ugly,FFM}.

>>>>

Linear Temporal Logic (LTL) is a common specification formalism for reactive and concurrent systems. It is often used in e.g. model checking and runtime verification. Another formalism that is used for the same purpose is finite automata, often over infinite words. This includes B\"{u}chi, Rabin, Street, Muller and Parity automata~\cite{Thomas}, all having the same expressive power. In fact, model checking of an LTL specification
is usually performed by first translating the specification into a B\"{u}chi automaton. The automata formalisms are more expressive than LTL, with a classical example By Wolper~\cite{Wolper} showing that
it is not possible to express in LTL that every even state in
the sequence satisfies $p$. This has motivated extending LTL
in various ways.
We are mainly motivated by runtime verification of executions {\em with data}, for which a first-order temporal logic is
appropriate. There, too, we show that expressiveness can be extended.

We present first an extension of  propositional LTL that is based on using additional (auxiliary) propositions, not appearing in the execution, which are dependent on the past 
of the current execution. These propositions obtain their value in a state as a function of the prefix of the execution up and including that state through a past temporal formula. This can be seen as adding
summary states to the temporal specification, resulting in a specification that is in between logic and finite automata representation. This extension, which we believe is conceptually simpler
and more intuitive than other suggested extensions, is shown to be
equivalent to the other common specification formalism:
B\"{u}chi automata, regular expressions and quantified LTL.

We then proceed to first-order LTL. We relativise Wolper's example to the first
order case to motivate the need for
extending first-order LTL. Another example is based on the in-expressiveness of transitive closure of a relation in first order logic, where, e.g., using a relation representing neighbors in a graph is not enough for representing that there is a path between nodes.
We present an extension where auxiliary relations, not appearing in the execution, are
added. While this extension
is capable of expressing the above
properties, we show that for the first-order version, this has less 
expressive power than adding alternating quantification over relations. This in distinction from adding quantification to propositional LTL, where prefixing existential quantification
provides the full expressive power of, say B\"{u}chi automata.

The extensions we proposed to the logics are in particular natural for
runtime verification of past (safety) temporal properties, since
the values of propositional and first-order LTL are based
on calculating and updating a summary of the prefix of the execution. Runtime verification is often restricted to a {\em past}  version of temporal logic~\cite{HR}, mainly because for past properties we can decide when the monitored property is violated immediately when this happens, while for future properties we may never know for sure. Consistent with that, 
we provide runtime verification algorithms for the past-time versions of the extended propositional and first-order
logics. We further present an extension of the 
\dejavu{} tool (publicly available on {\sf github},
hidden here because of anonymous reviewing)
\ignore{(https://github.com/havelund/dejavu)} that
realizes the extended first-order past temporal logic verification. The \dejavu{} tool~\cite{HPU,HP} allows runtime verification on past time first-order temporal logic over infinite domains (e.g., the integers, strings). It achieves efficiency by BDD representation of enumerations of data (rather than over the data values themselves) and using
a garbage collection algorithm.

The contributions of this paper are both theoretical and practical. On the theory side, we present and study extensions for propositional and first order linear-temporal logics and show the relations between them and existing versions. 
On the practical side, we also present RV algorithms for the extended logic, and 
show how to cast them specifically in an efficient BDD-based implementation. We present some experiments performed with our implementation and report on the results.

The structure of this paper is as follows. In Section~2 we first define the syntax and semantics of propositional temporal logic. We then review some of its classical extensions and present our own extension. In Section~3 we present a runtime verification algorithm for the extended past LTL. In Section~4 we review first order linear temporal logic, present our extension and study the relationships between the presented versions of first-order LTL. In Section~5 we present a runtime verification algorithm for the past version of the extended first-order LTL. Section~6 describes the implementation and provides experimental results. Finally, Section~7 concludes the paper.

@@

\vspace{1ex}
\noindent {\bf Related Work.}
For propositional LTL, the logic ETL~\cite{WVS,Wolper} extends
the set of temporal operators 
by using automata (or, equivalently a right linear grammar). Each automaton defines a temporal operator,
where the input letters of the automaton correspond to future temporal subformulas. Each accepted run defines then the places
where the subformulas (corresponding to the occurrences of the input letters in the run) need to hold. Another expressive-equivalent extension, 
in~\cite{Wolper}, adds dynamic, i.e.,
state-dependent, quantification over propositions to LTL.

@@

\begin{itemize}
\item{P} restricts to P{\em ast} temporal operators.
\item{F} allows F{\em irst-order} quantification over
data assigned to variables that is uniform for all states. This quantification is referred to as {\em static} quantification.
\item{Q} adds (second order) Q{\em uantification} that is dynamic (state-dependent). This is referred to as {\em dynamic} quantification.
\item{E} allows E{\em xtending} the 
set of objects that occur in the input with
new auxiliary predicates (for the propositional versions) or relations (for the first-order cases) using past temporal logic rules that define them.
\end{itemize}

@@

\section{Propositional LTL}
\label{propLTL}

The classical definition of linear temporal logic~\cite{MP} has the following syntax:
\[ \varphi ::= \true \, | \,  p \, | \, ( \varphi \wedge \varphi ) \, |  \, \neg  \varphi \, |   \, \bigcirc \varphi \, |  
\, ( \varphi \until \varphi ) \,|  \, \ominus \varphi |
   \, ( \varphi \since \psi ) \,  \]
where $p$ is a proposition from a finite set of propositions $P$, and $\bigcirc$, $\until$, $\ominus$, $\since$ stand for {\em next-time}, 
{\em until}, {\em previous-time} and {\em since}, respectively.

The models for LTL formulas 
are infinite sequence of states, of the form
$\sigma =  s_{1}\, s_{2}\, s_3 \ldots$,
where $s_i \subseteq P$ for each $i\geq 1$. These are the propositions that {\em hold} in that state.
%We denote by $\xi_{i}$ the suffix $s_{i} s_{i+1} s_{i+2}\ldots$ of $\sigma$. 
LTL's semantics is defined as follows:
\begin{itemize}
\item $( \sigma, {i}) \models \mathit{true}$.
\item $(\sigma, {i}) \models p$ iff $p \in s_{i}$.
\item $(\sigma, {i}) \models \neg\varphi$ iff  $(\sigma, {i}) \not\models \varphi$.
\item $(\sigma, {i}) \models ( \varphi \land \psi)$ iff $(\sigma, {i}) \models \varphi$ and $(\sigma, {i} ) \models \psi$.
\item $(\sigma, {i}) \models \bigcirc \varphi$ iff $(\sigma, {i+1}) \models \varphi$.
\item $(\sigma, {i}) \models ( \varphi \cal{U} \psi )$ iff for some $j$,
$j\geq i$, $(\sigma, {j}) \models \psi$,
and for each $k$, $i \leq k < j$, $(\sigma, {k}) \models \varphi$.
\item $(\sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
\item $(\sigma , i) \models (\varphi \since \psi)$ iff there exists $j$, $1 \leq j \leq i$, such that
$(\sigma , j)  \models \psi$ and for each
$k$, $j < k \leq i$, $(\sigma , k ) \models \varphi$.
\end{itemize}
Then $\sigma \models \varphi$ when $( \sigma ,1 ) \models \varphi$.
We can use the following abbreviations:
$\false = \neg \true$, 
$(\varphi \vee \psi) = \neg ( \neg \varphi \wedge \neg \psi )$, 
$(\varphi \rightarrow \psi) = ( \neg \varphi \vee \psi)$,
$(\varphi \leftrightarrow \psi) = ((\neg \varphi \wedge \neg \psi) \vee (\varphi \wedge \psi ))$,
$\Diamond \varphi = (\true \until \varphi)$ and $\Box \varphi = \neg \Diamond \neg \varphi$,
${\bf P}\  \varphi = (\true \; \since \; \varphi)$ ({\bf P} stands for P{\em reviously}), and 
${\bf H}\  \varphi = \neg {\bf P}\ \neg \varphi$ ({\bf H} stands for 
H{\em istory}).

%%infinite sequences (finite executions can be extended by repeating the last event indefinitely to form an infinite paths).
The expressive power of different versions of propositional LTL is often compared to regular expressions
over the alphabet $\Sigma = 2^P$ and to  {\em monadic} first and second order logics where the temporal propositions
correspond to unary predicates (over time). An overview appears in~\cite{Thomas}. Accordingly, we have the following characterizations:
LTL is equivalent to monadic first-order logic\footnote{First-order logic with unary predicates over the naturals, and the relation $<$.},  star-free regular expressions\footnote{Regular expressions but without star operator (or $\omega$).} and counter-free B\"{u}chi automata\footnote{A counter-free language is a regular language for which there is an integer $n$ such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}. In fact, restricting the temporal operators to the {\em future}  operators $\until$ and $\bigcirc$
(and the ones derived from them $\Box$ and $\Diamond$)
leaves the same expressive power and is commonly used.

@@

An important subset of LTL, called here PLTL, allows only past temporal
operators: $\since$, $\ominus$ and the operators derived
from them, $\mathbf{H}$ and $\mathbf{P}$. The past time logic is sometimes interpreted over finite sequences, 
where $\sigma \models \varphi$ when $( \sigma , | \sigma | ) \models \varphi$.
Alternatively,
it is also a common practice to use the past time restricted version, prefixed with  a single $\Box$ (always) operator; in this case, each of the prefixes has to satisfy $\varphi$. This later
form expresses {\em safety} LTL properties~\cite{AS} over infinite sequences. 
(When PLTL is interpreted over finite sequences, its
expressive power is the same as star-free regular expressions, first-order monadic logic over finite sequences and counting-free automata.)

@@

\subsubsection{Extending LTL with Dynamic Quantification.}
Adding quantification over propositions, as done in~\cite{Wolper}, allows writing a formula of the form
$\Exists q \, \varphi$, where
$\Exists q$ represents {\em existential} quantification over a proposition $q$ that can appear in $\varphi$.

@@

\begin{itemize}
\item $( \sigma, i ) \models \boldsymbol{\Exists} q \, \varphi$ iff there exists
$\sigma'$ such that $\sigma' |_{\{ q \} } = \sigma$ and 
$( \sigma', i) \models \varphi$.
\end{itemize}
%Since all the auxiliary variables appear in the scope of
%quantifiers, it holds that $\sigma \models \varphi$
%(in $\varphi$, auxiliary variables appear only within
%the scope of quantifiers)
%iff $\sigma |_A \models \varphi$.
%
%\klaus{Are the last two lines above needed?}
%
Universal quantification is also allowed, where $\Forall q\, \varphi =
\neg \Exists q\, \neg \varphi$.
This kind of quantification is considered to be {\em dynamic}, since the quantified variables can have different truth values depending on the states. It is also called  {\em second order} quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
Extending LTL with such quantification, the logic QLTL has the same expressive power as regular expressions, full B\"{u}chi
automata, or monadic second order logic with unary predicates over the naturals, see again~\cite{Thomas}.
In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of QLTL~\cite{Thomas}.

Wolper's property can be expressed in QLTL as follows:
\begin{equation} \label{form0} \Exists q \,
( (  \neg q \wedge \Box (q \leftrightarrow \bigcirc \neg q )  )  \wedge \Box (q \rightarrow p ) )\end{equation}

Restricting QPLTL to the past modalities, one obtain the logic PQLTL. PQLTL has the same expressive
power as regular expressions and finite automata.
Equation~(\ref{form0}) can
be rewritten in PQLTL as:
\begin{equation} \label{form1} \Exists q \,
{\bf H} (( q \leftrightarrow \ominus \neg q )   \wedge (q \rightarrow p ) )\end{equation}
%In fact, it can be shown that a single existential quantifier is sufficient. 
Since $\ominus \varphi $ is interpreted as  $\false$ in the first state of any sequence, regardless of $\varphi$, then
$q$ is $\false$ in the initial state. Then $q$ alternates between even and odd states.

%Note that since finite automata over finite sequences can be determinized, it is always sufficient to translate a
%PQLTL property into a deterministic
%automaton.

@@

We introduce now another extension of LTL, which we call ELTL.
%It has a form that puts it, conceptually, in between temporal
%logic and automata: on the one hand, it is possible to simply write
%a pure LTL formula, and on the other hand, one can write a specification that resembles a finite automaton. 
As will be showed later, this extension is
is very natural for runtime verification.

We partition the propositions $P$ into
{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
and {\em basic propositions} $B$.
%A model for an ELTL formula has truth assignments only over the propositions in $B$.
An ELTL property $\eta$ has the following form: 
\begin{equation} \label{ELTL}
\psi \mathit{\ where\ } a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}} \end{equation}
where each $a_j$ is a distinct auxiliary proposition from $A$,
$\psi$ is an LTL property and each $\varphi_i$ 
is a PLTL property where propositions from $B$ can
only occur within the scope of a $\ominus$ operator.
The semantics can be defined as follows. We refer to $\psi$
as the {\em statement} of $\eta$ and to 
$a_j  := \varphi_j$ as a {\em rule}. In text, rules
 will be separated from each other by commas. 
\begin{tabbing}
\ \ \ $\sigma \models \eta$ \=  if there exists $\sigma'$, where
$\sigma' |_A = \sigma$ such that \\
 \> $\sigma' \models ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j)) $
\end{tabbing}

\noindent
Wolper's example can be written in ELTL
as follows:
\begin{equation} 
\label{form3}
\Box ( q \rightarrow p ) \, \mathrm{where} \, 
 ( q := \ominus \neg q)
\end{equation}
where $A = \{ q\}$ and $B = \{ p \}$.
The auxiliary variable $q$ is used to augment the input sequence such that each {\em odd} state will satisfy $\neg q$
and each {\em even} state will satisfy $p$.

ELTL extends the set of propositional variables with new variables, whose values at a state are
functions of (i.e., uniquely defined by) the prefix of the model up to that state. The added propositions effectively summarize the past execution prefix up to and
including the current state.
This differs from 
the use of auxiliary variables in QLTL, where the values assigned to the
auxiliary variables do not have to extend the 
states of the model in a unique way throughout the interpretation of 
the property over a model. The constraint that auxiliary variables appearing in the formulas $\varphi_i$ 
must occur within the scope of a $\ominus$ operator
is required to help preventing that the values
of the auxiliary variables are conflicting, as in
$a_1 := \neg a_2$ and $a_2 := a_1$, which makes a formula
that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write conflicting rules, e.g., by using the rule $a_1 : = \false$.

@@

\noindent {\bf Sketch of proof.} By induction on the state index. Each auxiliary variable
is defined, via a PLTL formula by the current state of the model over the values of the propositions $B$, and by the values of $A \cup B$ in previous states.  \qed
%A PLTL property depends in a state only on the prefix of the model that precedes and includes that state.

\begin{theorem}
\label{theo1}
The expressive power of 
ELTL is the same as QLTL.
\end{theorem}
\noindent {\bf Proof.}
Each ELTL formula $\eta$, as defined above,
is expressible using the following equivalent QLTL formula:

@@

\noindent
For the other direction, one can translate the QLTL property first into
a second order monadic logic formula, then to a deterministic Muller automata and then construct an ELTL formula that holds for the accepting executions of this automaton. The rules of this formula encode the automata states, and the statement describes the acceptance condition of the Muller automaton.
The details of the
constructions are given in the appendix. \qed

@@

We define EPLTL by disallowing the future temporal operators (except an implied prefixing $\Box$) in ELTL. This results in
a formalism that is equivalent to finite automata, where all the states except one are accepting and where the non-accepting state is a sink. 
We can use a related, but simpler construction than in Theorem~\ref{theo1} to prove the following:

@@

\begin{lemma}
\label{simplecase}
The expressive power of 
EPLTL is the same as QPLTL.
%Any safety QPLTL property can be expressed using the following restricted
%ELTL form:
%\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
% {j \in \{1, \ldots , n\}}  \] 
% with $p$ being one of the auxiliary variables %$a_j$.
\end{lemma}
In fact, it is easy to show that EPLTL can be restricted to the following form, while retaining its expressive power.
\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
{j \in \{1, \ldots , n\}}  \] 
with $p$ being one of the auxiliary variables $a_j$. Yet, this may not be the preferred form
of an EPLTL property, and the use of further past-time temporal operators, and a non trivial statement  part, may provide a more intuitive specification.
The relative expressive power between the propositional temporal logics discussed in this section is summarized in Figure~\ref{prop}. An arrow from A to B means that logic  B is more expressive than A. A double headed arrow shows same expressiveness

@@

\begin{figure}
\begin{center}
%\includegraphics[height=0.6in,width=2in]{PROP.jpg}
\includegraphics[width=2.8in]{PROP.jpg}
\caption{\label{prop}
Comparing the expressive power of the temporal logics in Section~\ref{propLTL}
}
\end{center}
\end{figure}

@@

Runtime verification of temporal specifications often
concentrates on the past portion of the logic.  Past-time  specifications have the important property that one can distinguish when they are violated after observing a finite prefix of
an execution. For an extended discussion of this issue of {\em monitorability}, see e.g.,~\cite{Ugly,FFM}.

The algorithm for PLTL, first presented in \cite{HR}, 
is based on the observation that the semantics of the 
past time formulas $\ominus \varphi$ and $(\varphi\, \since \,\psi)$ in the current step $i$ is defined in terms of the semantics
in the previous step $i - 1$ of a subformula.
To demonstrate this, we repeat here the definition of the previous-time
operator $\ominus$, and rewrite the definition
of $\since$ in an equivalent form that is
more directly applicable for runtime verification.
\begin{itemize}
\item $( \sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
\item $(\sigma , i) \models (\varphi \, \since \, \psi)$ iff $(\sigma , i) \models \psi$ or the following hold: $i>1$,
$( \sigma , i)  \models \varphi$ and 
$(\sigma , i-1 ) \models ( \varphi \, \since \, \psi )$.
\end{itemize}
Runtime verification for a PLTL formula $\eta$ exploits the fact that
the semantic definition is recursive in both the structure of the property. Thus, subformulas
are evaluated based on smaller subformulas, and the
evaluation of subformulas in the previous state.
The algorithm, shown below, uses two vectors (arrays) of values indexed by subformulas:  $\old$, which summarizes the truth values of the
subformulas for the execution prefix that
ends just {\em before} the new state, and $\current$, for the execution prefix that
ends with the {\em current} state. The order of calculating $\current$ for subformulas is bottom up, according to the syntax tree.
\begin{enumerate}
\item Initially, for each subformula $\varphi$
of $\eta$,
%$B ( \varphi , \current ) = \bfalse $.
$\current ( \varphi ) := \false$.
